Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-07-16 Thesis/Background.tex
Semi-final presubmission draft

\chapter{A Dynamical Recipe for Cosmological Disks}\label{ch:background}

This chapter provides an overview of the background theory needed to understand subsequent chapters. In Chapter~\ref{ch:introduction}, we had a lot of discussion about simulations of galaxies being critical to having a theoretical understanding of cosmology and galaxy formation. The word ``simulation" was used without fully contextualizing its significance and meaning. We provide that context in \S\ref{sec:motivation}, where we describe what is actually being done when a simulation is run. In \S\ref{sec:galaxy_ics}, we talk about how to apply this theory to study the evolution of equilibrium galaxies. We also spoke at length about $\Lambda$CDM cosmology and how it is critical to explain the evolution of galaxies on a global scale. We talk about cosmology in detail in \S\ref{sec:cosmology}, and explain how this specifies an extended view of the discussion in \S\ref{sec:motivation}. Lastly, we talk about various techniques relied upon in the analysis of simulation in \S\ref{sec:analysis_of_sims}. This includes how we identify cosmic substructure and the techniques we use for analyzing the time-evolution of disk-derived quantities. This chapter summarizes the very basics of the models used in this thesis. It is by no means a comprehensive account of galactic dynamics. In the course of discussion, we will point the reader to more detailed accounts of the topics being discussed.

\section{Physical Motivation of Modeling} \label{sec:motivation}

@@

Unlike gas, which can exhange thermodynamic energy with itself through a variety of mechanisms, stars and dark matter interact only through gravity. Gravity is a long range force. In fact, the majority of the contribution to the forces on stars in galaxies comes from far outside their immediate neighborhoods (that is to say gravitational energy in non-extensive) \citep{BT}.

This has a number of interesting implications. Classical equilibrium statistical mechanics tries to understand distributions of particles as ensembles with a given thermodynamic potential. Some commonly used ensembles are the microcanonical ensemble, where $f_0(\textbf{r},\textbf{v})$, the equilibrium state, is determined by holding energy fixed in a closed system, and the canonical ensemble which finds $f_0$ holding the temperature of the system fixed \citep{sethna}. These ensembles break down when non-extensive forces are involed (in the latter case, specifically because self-gravitating systems have no maximum entropy) \citep{self_gravitating_statistical_mechanics,lb_negative_specific_heat, BT}. Another peculiarity of self-gravitating systems is that they have a negative heat capacity \citep{lb_negative_specific_heat}.

The implication is that self-gravitating systems of stars, which have no means to dissipate internal energy, can never be in thermodynamic equilibrium. A system which starts in dynamical equilibrium evolves to a state of higher entropy when perturbed, which is physically characterized as having a dense core and extended envelope of mass \citep{BT}. Since a thermodynamic resolution of why galaxies evolve the way the do is unattainable due to the extensive nature of gravity, we are forced to study non-equilibrium dynamical evolution of $f(\textbf{r},\textbf{v},t)$ to understand galactic systems, and this requires more theory\footnote{For more information on the thermodynamics of self-gravitating systems, Chapter~4.10 of \citet{BT} and Chapter~7.3 of \citet{BT} provide excellent summaries.}.

@@

In terms of actually solving the CBE, we run into a number of technical challenges in applying traditional finite difference and finite volume schemes for advection equations. These techniques require that some mesh be constructed, and the flows between the cells in the mesh are calculated dependent on the type of equation being solved and the current states of the cells\footnote{Chapter~19 of \citet{numerical_recipes_fortran} discusses numerical methods for PDEs, although you could find detailed discussion of finite difference and finite volume methods in most numerical methods texts.}. The spatial domain is 6D, meaning a uniform mesh resolution scales in memory consuption as $O(N^6)$, where $N$ is the number of cells on an axis. For a sanity check, a modest $N = 2^5$ element grid would take up $2^{30} \times 2^4 \text{ bytes} \sim 17 \text{ GB}$. We obviously need more than 32 cells on an axis to represent the fine DF structure we want to study, and we are already at the limit of a single compute node to apply a finite difference scheme.

@@

The point of introducing this powerful concept is simply in recognition of the fact that we need to integrate over phase space to solve Eq. \eqref{eq:cbe_cartesian}. As we stated, computing these integrals on a mesh is intractible, so we represent $f(\textbf{r},\textbf{v},t)$ in another way:
\begin{equation}
f(\textbf{r},\textbf{v},t) \approx \sum_i m_i \delta^3(\textbf{r} - \textbf{r}_i) \delta^3(\textbf{v} - \textbf{v}_i)
\end{equation}
where $\delta^3$ is the 3D Dirac delta function, and we take the DF to be normalized to $M = \sum_i m_i$ instead of 1. We call this the N-body realization of the DF. As an extension of this idea, to find $f(\textbf{r},\textbf{v},t)$ at some point in the future, it is sufficient to find the state of our \textit{approximation}\footnote{The author believes that this is one of the most beautiful results ever to be applied in studying the dynamics of galaxies.} at some time in the future. This requires a more detailed exploration of how we step through time with N-body realizations.

@@

The final remaining complication for the timestepping portion of solving the CBE is determining the timesteps themselves. This is slightly more complicated to do properly than choosing a fixed timestep for all particles. This is because accelerations and velocities in astrophysical systems vary by orders of magnitude, and the timestep needs to be small compared to the timescales defined by the relevant accelerations. The way that this is commonly implemented, including in \citet{GadgetCodePaper}, is to have a base timestep for all particles that gets bisected at each level of refinement. That is to say, particles at level 4 are updated four more times than particles at level 2. Suppose the highest level of refinement is $k$. The simulation proceeds in $\Delta t_{base} / 2^{k}$ intervals, with particles at $k - 1$ being updated at half of the timesteps, dividing the number of updates by two up to the coarsest level of temporal refinement. Where multiple levels need updates, particles at the lower levels are updated first.

The level of temporal refinement assigned to a given particle is determined primarily by the acceleration it experiences. That is, for \textsc{Gadget-3} \citep{GadgetCodePaper},
\begin{equation}
\Delta t_i = \min\left(\Delta t_{base}, \left(\frac{2 \eta \epsilon}{\vert \nabla \Phi(\textbf{r}_i) \vert} \right)^{1/2}\right).
\end{equation}
Here, $\eta$ is a free accuracy parameter and $\epsilon$ is the gravitational softening. The gravitational softening is used to prevent numerical overflow when particles rarely have close encounters. It arises from having the acceleration induced from a particle at position $\textbf{r}_i$  be,
\begin{equation}
\textbf{a}_i(\textbf{r}) = -\frac{G m_i}{\left(\vert\textbf{r} - \textbf{r}_i\vert^2 + \epsilon^2\right)^{3/2}}(\textbf{r} - \textbf{r}_i) \label{eq:point_mass}
\end{equation}
Higher choices of $\epsilon$ reduce large errors accumulating due to unsustainably large forces, but also make the force calculation less accurate.

@@

The Leapfrog scheme requires us to compute the force two times. Naively, we would compute a sum over Eq. \eqref{eq:point_mass} for each of the $N(N-1)/2$ unique pairs of positions in $O(N^2)$ time to compute the forces on all particles. This is also intractible, or rather is the limiting factor on how large of a simulation we can run. Methods were developed early on in the history of running N-body simulations to cope with this by approximating the potential.

The first method worth noting for this thesis is the particle mesh technique \citep{hockney_pm, white_pm, klypin_pm}. This technique considers the calculation of forces on a 3D mesh by reducing the calculation to a Fast Fourier Transform (FFT) \citep{numerical_recipes_fortran}. One of the crowning achievements of scientific computing in the 20th century was the rediscovery of an algorithm by J. W. Cooley and John Tukey, originally discovered by Gauss, to compute the discrete fourier transform (DFT) in $O(N\,\log\,N)$ complexity\footnote{What Gauss was doing with recursive DFT algorithms is beyond the author, but in the author's humble opinion, it is expected from the greatest mathematician to ever live.}\citep{fft}. Particle mesh's cloud-in-cell (CIC) algorithm was a part of a flurry of algorithms that subsequently exploited the FFT. It has a difficult time accounting for forces from particles separated by small scales on the mesh, though \citep{GadgetCodePaper}.

Compensating for some of the shortcomings of the particle mesh approach, \citet{barnes_hut} introduced a new algorithm for approximating forces in $O(N \,\log\, N)$ complexity. Their algorithm took a divide an conquer approach by subdividing the simulation space into octants recursively until each particle was in its own cell. Forces would be computed pairwise for particles close together, but for distant particles, the mass distrubtion would be approximated as a multipole. To compute the forces, one simply does a depth-first-search (DFS) of the particle tree, going deeper if,
\begin{equation}
\frac{l}{d} \geq \theta
\end{equation}
where $l$ is the side length of the node whose force is being computed, $d$ is the distance from this node to the current node in the DFS, and $\theta$ is a threshold called the opening angle. In the simplest case, a cell not meeting this condition would be approximated as a point mass with the position at the  center of mass of all particles deeper in its subtree. Higher order multipole expansions are typically used (4 and 8 are common choices).

@@

Before proceeding to talk about the complications of executing TreePM on distributed systems like the Linux clusters worked with in this thesis, we would be remiss in not mentioning field expansion methods. Typically these take the form of an expansion of the potential in spherical harmonics using a set of radial basis functions. A brief discussion of these methods is presented in Chapter~2.9.4 of \citet{BT}. These methdos have great historical significance, and have even seen a recent resurgence in helping understand the structure of dark matter halos \citep{lilley_2018_a, lilley_2018_b}.

To summarize, there are a number of ways to find gravitational forces in a simulation that all amount to solving Poisson's Equation, \eqref{eq:poisson}. The most accurate way is to compute all pairwise forces, but efficient algorithms have been developed to accurately estimate forces from gravity. The force calculation together with a timestepping scheme for N-body realizations forms the main components of an N-body simulation, and solving Eq. \eqref{eq:nbody_system} to estimate $f(\textbf{r},\textbf{v},t)$ is what we mean when we say we ran a simulation.

@@

To reduce the need to explicitly transfer particles between processes, \textsc{Gadget-3} notes that pairwise forces will only be needed for nearby poarticles. These represent the bulk of the tree force calculation. Any scheme which puts nearby particles on the same processor will reduce the overall communication time in the force calculation phase. One way to accomplish efficient load balancing is to partition the simulation space using a space filling curve. Fig. \ref{fig:peano} shows the Peano space-filling curve at several levels of refinement in 2D. This is the curve \textsc{Gadget-3} uses. A level of refinement is selected, and estimates of the computational cost in each cube are added up along the space filling curve to equally distribute the computational load. This results in nearby particles being on the same process, and in the processes having roughly equal work loads. \citet{GadgetCodePaper} shows that this is highly effective at reducing communication cost in the simulation, and it goes into more detail about how TreePM is carried out on distributed systems.

@@

\subsection{Jeans Modeling and the Epicyclic Approximation}

@@

However, the DF is not the quantity that is observed in external galaxies. Astronomers are not typically working with a 6D phase space description of the structures they observe. Luminosity (density) profiles are one observable , as well as circular velocities (rotation curves) for disks, and also the spread of line of sight (LOS) velocties of disk stars. A modelling approach that takes our understanding of these observables as input is conceptually easier to understand than a DF which may or may not be unique \citep{BT}.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} was one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, disks, and dark matter halos. In the case of the dark matter halo for a disk-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} proceeded this work in establishing a common density profile for dark matter halos). As such, we will not talk about their uninformed choice of halo, and it really is not important.

In the case of the axisymmetric disk, \citet{hernquist_1993} based their model on the observations at the time which suggested that stellar disks had an exponential radial density profile \citep{freeman_1970}. This motivated a density profile commonly used today,
\begin{equation}
\rho_d(r) = \frac{M_d}{4 \pi R_d^2 z_d} e^{-R/R_d} \text{sech}^2\left(\frac{z}{z_d} \right). 
\end{equation}
Here, $M_d$ is the mass of the disk, $R_d$ is the radial scale length, and $z_d$ is the vertical scale length.
Observations support the idea that the second radial velocity moment also has an exponential profile \citep{van_der_kruit_searle_1981, lewis_freeman_1989},
\begin{equation}
\langle v_R^2  = \sigma_{R,0} e^{-R/R_\sigma}.
\end{equation}
Here, $R_\sigma$ is the radial dispersion scale length, and $\sigma_{R,0}$ is a free velocity scale (on the order of 100 km/s for the Milky Way). In  priciple, any radial scale length can be used for $\langle v_R^2 \rangle$, but it is generally accepted to be a longer scale length than $R_d$. The vertical dispersion and second velocity moment are determined from the assumption of the disc being isothermal,
\begin{equation}
\langle v_z^2 \rangle = \pi G \Sigma z_d
\end{equation}
where $\Sigma_d$ is the surface density as a function of radius. Finally, a relation between $\langle v_\phi^2 \rangle$ and the other second moments is needed to fully solve the second order system. Jeans modelling for discs injects the intuition behind a theoretical development known as the \textit{epicycle approximation}.

The main idea is that stars in discs are on roughly circular and planar orbits, so a reasonably accurate description of their orbits can be obtained by taking the equations of motion in a truncated series expansion about a circular orbit at the guiding radius, $R_g$. This is better understood in terms of the effective potential in the rotating reference frame in which a star rotating at $R_g$ is stationary. For a star on the $x$-axis, the $\hat{x}$ direction represents the radial direction, the $\hat{y}$ direction is the same as $\hat{\phi}$ in the original frame, and $\hat{z}$ is unchanged. The equations of motion are \citep{BT},
\begin{eqnarray}
\frac{\text{d}^2x}{\text{d} t^2} &=& \deriv{\Phi_{eff}}{R}\\
\frac{\text{d}^2y}{\text{d} t^2} &=& 0\\
\frac{\text{d}^2z}{\text{d} t^2} &=& \deriv{\Phi_{eff}}{z}
\end{eqnarray}
where $\Phi_{eff} = \Phi(R,z) + L_z^2/2 R^2$ is the efffective potential in the accelerating (rotating) reference frame, and $L_z$ is the $z$-component of the angular momentum. These equations can be written in a first order expansion as,
\begin{eqnarray}
\frac{\text{d}^2x}{\text{d} t^2} &=& \kappa^2 x\\
\frac{\text{d}^2y}{\text{d} t^2} &=& 0\\
\frac{\text{d}^2z}{\text{d} t^2} &=& \nu^2 z
\end{eqnarray}
where,
\begin{equation}
\kappa^2(R_g) = \left(R \deriv{\Omega^2}{R} + 4 \Omega^2\right)_{R=R_g,z=0},
\end{equation}
with the circular frequency $\Omega$,
\begin{equation}
\Omega^2(R_g) = \frac{1}{R_g} \left(\deriv{\Phi}{R}\right)_{R=R_g,z=0} = \frac{L_z^2}{R_g^4},
\end{equation}
and,
\begin{equation}
\nu^2(R_g) = \left(\deriv{\Phi}{z}\right)_{R=R_g,z=0}.
\end{equation}
That is, a star's orbit is decribed by three frequencies: its rotational frequency $\Omega$, a radial oscillation frequency, $\kappa$, and a vertical oscillation frequency, $\nu$. This view is accurate so long as the vertical motion does not deviate far from the disc midplane. The actual values of these constants in Galactic astonomy are related to a problem known as the \textit{Oort Problem}, and we believe that for the Sun \citep{BT},
\begin{eqnarray}
\frac{\kappa}{\Omega} &\approx& 1.35 \\
\frac{\nu}{\kappa} &\approx& 2
\end{eqnarray} 
The fact that the orbit is described by three distinct frequencies that are unequal for the Sun means that the Sun's orbit does not close. If the Milky Way potential never changed, the path of the Sun would fill an annulus centered around $R_g$ \citep{BT}. That is to say, the orbit looks like it forms a rosette pattern.

@@

Although Jeans modeling is used to set up initial conditions for galaxies because it does not require a DF to be known ahead of time, the best initial conditions are generated from DFs \footnote{That is not to say that no one uses Jeans modelling to set up ICs. Approaches like the one presented in \citet{hernquist_1993} were enormously successful and still used today.}. In general, the equilibrium DF is a function of six variables. However, there are some quantities, called \textit{integrals of motion}, which do not change along a star's orbit in a potential. For an axisymmetric potential, these are the total energy, $\mathcal{E}$, $L_z$ and a third quantity, $I_3$.  
For spherical systems, up to five integrals of motion are admitted by the system, but the simplest \textit{ergodic} models simply have $f_{h}(\mathcal{E})$ for the DF.

If we assume the epicycle approximation, then $I_3 \approx E_z = \Phi(R,0) - \Phi(R,z) + \frac{1}{2} v^2$, where $v$ is the velocity of an orbit \citep{kd95galactics}. The vertical energy, $E_z$, is conserved so long as the disc is razor thin. Jeans showed that a DF that is a function of the integrals of motion describes an equilibrium distribution. More specifically, quoting from \citet{BT},
\begin{theorem}[The Jeans Theorem]
Any steady-state solution of the collisionless Boltzmann equation depends on the phase-space coordinates only through integrals of motion in a givcen potential, and any function of the integrals yields a steady-state solution of the collisionless Boltzmann equation.
\end{theorem}
Proof of this theorem can be obtained in \citet{jeans_1915} and in Chapter~4.2 of \citet{BT}, although it essentially amounts to observing that the time derivative of a function of time-independent quantities is zero by the chain rule. Now that we know that any DF of the intergrals of motion gives a steady state solution, and that using the epicyclic approximation gives us three known integrals of motion, we can ask ourselves what kinds of DFs describe the NFW haloes and exponential disks that are easily obtained from Jeans modelling.

>>>>

\chapter{A Dynamical Recipe for Cosmological Disks}\label{ch:background}
\newpage

This chapter provides an overview of the background theory needed to understand subsequent chapters. In Chapter~\ref{ch:introduction}, we discussed simulations of galaxies being critical to having a theoretical understanding of cosmology and galaxy formation. The word ``simulation" was used without fully contextualizing its significance and meaning. We provide that context in \S\ref{sec:motivation}, where we describe what is actually being done when a simulation is run. In \S\ref{sec:galaxy_ics}, we talk about how to apply this theory to study the evolution of equilibrium galaxies.  We talk about  the implications of cosmology in detail in \S\ref{sec:cosmology}, and explain how this specifies an extended view of the discussion in \S\ref{sec:motivation}. This includes how we identify cosmic substructure and the techniques we use for analyzing the time-evolution of disk-derived quantities. 
\section{Physical Motivation of Modelling} \label{sec:motivation}

@@

Unlike gas, which can exchange thermodynamic energy with itself through a variety of mechanisms, stars and dark matter interact only through gravity. Gravity is a long range force. In fact, the majority of the contribution to the forces on stars in galaxies comes from far outside their immediate neighborhoods (that is to say gravitational energy in non-extensive) \citep{BT}.

This has a number of interesting implications. Classical equilibrium statistical mechanics tries to understand distributions of particles as ensembles with a given thermodynamic potential. Some commonly used ensembles are the microcanonical ensemble, where $f_0(\textbf{r},\textbf{v})$, the equilibrium state, is determined by holding energy fixed in a closed system, and the canonical ensemble which finds $f_0$ holding the temperature of the system fixed \citep{sethna}. These ensembles break down when non-extensive forces are involved (in the latter case, specifically because self-gravitating systems have no maximum entropy) \citep{self_gravitating_statistical_mechanics,lb_negative_specific_heat, BT}. Another peculiarity of self-gravitating systems is that they have a negative heat capacity \citep{lb_negative_specific_heat}.

The implication is that self-gravitating systems of stars, which have no means to dissipate internal energy, can never be in thermodynamic equilibrium. A system which starts in dynamical equilibrium evolves to a state of higher entropy when perturbed, which is physically characterized as having a dense core and extended envelope of mass \citep{BT}. Since a thermodynamic resolution of galactic dynamics is unattainable, a dynamical theory of galaxy evolution must be developed\footnote{For more information on the thermodynamics of self-gravitating systems, Chapter~4.10 of \citet{BT} and Chapter~7.3 of \citet{BT} provide excellent summaries.}.

@@

In terms of actually solving the CBE, we run into a number of technical challenges in applying traditional finite difference and finite volume schemes for advection equations. These techniques require that some mesh be constructed, and the flows between the cells in the mesh are calculated dependent on the type of equation being solved and the current states of the cells\footnote{Chapter~19 of \citet{numerical_recipes_fortran} discusses numerical methods for PDEs, although you could find detailed discussion of finite difference and finite volume methods in most numerical methods texts.}. The spatial domain is 6D, meaning a uniform mesh resolution scales in memory consumption as $O(N^6)$, where $N$ is the number of cells on an axis. For a sanity check, a modest $N = 2^5$ element grid would take up $2^{30} \times 2^4 \text{ bytes} \sim 17 \text{ GB}$. We obviously need more than 32 cells on an axis to represent the fine DF structure we want to study, and we are already at the limit of a single compute node to apply a finite difference scheme.

@@

The point of introducing this powerful concept is simply in recognition of the fact that we need to integrate over phase space to solve Eq. \eqref{eq:cbe_cartesian}. As we stated, computing these integrals on a mesh is intractable, so we represent $f(\textbf{r},\textbf{v},t)$ in another way:
\begin{equation}
f(\textbf{r},\textbf{v},t) \approx \sum_i m_i \delta^3(\textbf{r} - \textbf{r}_i) \delta^3(\textbf{v} - \textbf{v}_i)
\end{equation}
where $\delta^3$ is the 3D Dirac delta function, and we take the DF to be normalized to $M = \sum_i m_i$ instead of 1. We call this the N-body realization of the DF. As an extension of this idea, to find $f(\textbf{r},\textbf{v},t)$ at some point in the future, it is sufficient to find the state of our \textit{approximation}\footnote{The author believes that this is one of the most beautiful results ever to be applied in studying the dynamics of galaxies.} at some time in the future. This requires a more detailed exploration of how we step through time with N-body realizations.

@@

The final remaining complication for the time stepping portion of solving the CBE is determining the timesteps themselves. This is slightly more complicated to do properly than choosing a fixed timestep for all particles. This is because accelerations and velocities in astrophysical systems vary by orders of magnitude, and the timestep needs to be small compared to the timescales defined by the relevant accelerations. The way that this is commonly implemented, including in \citet{GadgetCodePaper}, is to have a base timestep for all particles that gets bisected at each level of refinement. That is to say, particles at level 4 are updated four more times than particles at level 2. Suppose the highest level of refinement is $k$. The simulation proceeds in $\Delta t_{base} / 2^{k}$ intervals, with particles at $k - 1$ being updated at half of the timesteps, dividing the number of updates by two up to the coarsest level of temporal refinement. Where multiple levels need updates, particles at the lower levels are updated first.

The level of temporal refinement assigned to a given particle is determined primarily by the acceleration it experiences. That is, for \textsc{Gadget-3} \citep{GadgetCodePaper},
\begin{equation}
\Delta t_i = \min\left(\Delta t_{base}, \left(\frac{2 \eta \epsilon}{\vert \nabla \Phi(\textbf{r}_i) \vert} \right)^{1/2}\right).
\end{equation}
Here, $\eta$ is a free accuracy parameter and $\epsilon$ is the gravitational softening. The gravitational softening is used to prevent numerical overflow when particles rarely have close encounters. It arises from having the acceleration induced from a particle at position $\textbf{r}_i$  be,
\begin{equation}
\textbf{a}_i(\textbf{r}) = -\frac{G m_i}{\left(\vert\textbf{r} - \textbf{r}_i\vert^2 + \epsilon^2\right)^{3/2}}(\textbf{r} - \textbf{r}_i) \label{eq:point_mass}
\end{equation}
Higher choices of $\epsilon$ reduce large errors accumulating due to large forces, but also make the force calculation less accurate.

@@

The Leapfrog scheme requires us to compute the force two times. Naively, we would compute a sum over Eq. \eqref{eq:point_mass} for each of the $N(N-1)/2$ unique pairs of positions in $O(N^2)$ time to compute the forces on all particles. This is also intractable, or rather is the limiting factor on how large of a simulation we can run. Methods were developed early on in the history of running N-body simulations to cope with this by approximating the potential.

The first method worth noting for this thesis is the particle mesh technique \citep{hockney_pm, white_pm, klypin_pm}. This technique considers the calculation of forces on a 3D mesh by reducing the calculation to a Fast Fourier Transform (FFT) \citep{numerical_recipes_fortran}. One of the crowning achievements of scientific computing in the 20th century was the rediscovery of an algorithm by J. W. Cooley and John Tukey, originally discovered by Gauss, to compute the Discrete Fourier Transform (DFT) in $O(N\,\log\,N)$ complexity\footnote{What Gauss was doing with recursive DFT algorithms is beyond the author, but in the author's humble opinion, it is expected from the greatest mathematician to ever live.}\citep{fft}. Particle mesh's cloud-in-cell (CIC) algorithm was a part of a flurry of algorithms that subsequently exploited the FFT. It has a difficult time accounting for forces from particles separated by small scales on the mesh, though \citep{GadgetCodePaper}.

Compensating for some of the shortcomings of the particle mesh approach, \citet{barnes_hut} introduced a new algorithm for approximating forces in $O(N \,\log\, N)$ complexity. Their algorithm took a divide an conquer approach by subdividing the simulation space into octants recursively until each particle was in its own cell. Forces would be computed pairwise for particles close together, but for distant particles, the mass distribution would be approximated as a multipole. To compute the forces, one simply does a depth-first-search (DFS) of the particle tree, going deeper if,
\begin{equation}
\frac{l}{d} \geq \theta
\end{equation}
where $l$ is the side length of the node whose force is being computed, $d$ is the distance from this node to the current node in the DFS, and $\theta$ is a threshold called the opening angle. In the simplest case, a cell not meeting this condition would be approximated as a point mass with the position at the  center of mass of all particles deeper in its subtree. Higher order multipole expansions are typically used (4 and 8 are common choices).

@@

Before proceeding to talk about the complications of executing TreePM on distributed systems like the Linux clusters worked with in this thesis, we would be remiss in not mentioning field expansion methods. Typically these take the form of an expansion of the potential in spherical harmonics using a set of radial basis functions. A brief discussion of these methods is presented in Chapter~2.9.4 of \citet{BT}. These methods have great historical significance, and have even seen a recent resurgence in helping understand the structure of dark matter halos \citep{lilley_2018_a, lilley_2018_b}.

To summarize, there are a number of ways to find gravitational forces in a simulation that all amount to solving Poisson's Equation, \eqref{eq:poisson}. The most accurate way is to compute all pairwise forces, but efficient algorithms have been developed to accurately estimate forces from gravity. The force calculation together with a time stepping scheme for N-body realizations forms the main components of an N-body simulation, and solving Eq. \eqref{eq:nbody_system} to estimate $f(\textbf{r},\textbf{v},t)$ is what we mean when we say we ran a simulation.

@@

To reduce the need to explicitly transfer particles between processes, \textsc{Gadget-3} notes that pairwise forces will only be needed for nearby particles. These represent the bulk of the tree force calculation. Any scheme which puts nearby particles on the same processor will reduce the overall communication time in the force calculation phase. One way to accomplish efficient load balancing is to partition the simulation space using a space filling curve. Fig. \ref{fig:peano} shows the Peano space-filling curve at several levels of refinement in 2D. This is the curve \textsc{Gadget-3} uses. A level of refinement is selected, and estimates of the computational cost in each cube are added up along the space filling curve to equally distribute the computational load. This results in nearby particles being on the same process, and in the processes having roughly equal work loads. \citet{GadgetCodePaper} shows that this is highly effective at reducing communication cost in the simulation, and it goes into more detail about how TreePM is carried out on distributed systems.

@@

\subsection{Jeans Modelling and the Epicyclic Approximation}

@@

However, the DF is not the quantity that is observed in external galaxies. Astronomers are not typically working with a 6D phase space description of the structures they observe. Luminosity (density) profiles are one observable , as well as circular velocities (rotation curves) for disks, and also the spread of line of sight (LOS) velocities of disk stars. A modelling approach that takes our understanding of these observables as input is conceptually easier to understand than a DF which may or may not be unique \citep{BT}.

@@

Despite these shortcomings, more information can be imposed on the system to make these equations applicable. \citet{hernquist_1993} presented one of the first successful applications of Jeans modelling to creating equilibrium galaxies with bulges, disks, and dark matter halos. In the case of the dark matter halo for a disk-halo system, one only has to specify the function,
\begin{equation}
\beta(r) = 1 - \frac{\langle v_\theta^2\rangle + \langle v_\phi^2 \rangle}{2 \langle v_R^2 \rangle}.
\end{equation}
This parameter measures anisotropy in the velocity ellipsoid defined by $\sigma_{ij}$ at each radius; $\beta = -\infty$ if all orbits are circular, $\beta=0$ if the system is isotropic, and $\beta = 1$ if orbits are purely radial. Together with an assumed density-potential pair, continuity equation, and the spherical second-order Jeans equation, the spherical halo's velocity structure is specified. At the time of writing, \citet{hernquist_1993} did not have much information about the dark matter distribution in galaxies (recall \citep{nfw} proceeded this work in establishing a common density profile for dark matter halos). As such, we will not talk about their uninformed choice of halo, and it really is not important.

In the case of the axisymmetric disk, \citet{hernquist_1993} based their model on the observations at the time which suggested that stellar disks had an exponential radial density profile \citep{freeman_1970}. This motivated a density profile commonly used today,
\begin{equation}
\rho_d(r) = \frac{M_d}{4 \pi R_d^2 z_d} e^{-R/R_d} \text{sech}^2\left(\frac{z}{z_d} \right). 
\end{equation}
Here, $M_d$ is the mass of the disk, $R_d$ is the radial scale length, and $z_d$ is the vertical scale length.
Observations support the idea that the second radial velocity moment also has an exponential profile \citep{van_der_kruit_searle_1981, lewis_freeman_1989},
\begin{equation}
\langle v_R^2 \rangle = \sigma_{R,0} e^{-R/R_\sigma}.
\end{equation}
Here, $R_\sigma$ is the radial dispersion scale length, and $\sigma_{R,0}$ is a free velocity scale (on the order of 100 km/s for the Milky Way). In  principle, any radial scale length can be used for $\langle v_R^2 \rangle$, but it is generally accepted to be a longer scale length than $R_d$. The vertical dispersion and second velocity moment are determined from the assumption of the disc being isothermal,
\begin{equation}
\langle v_z^2 \rangle = \pi G \Sigma z_d
\end{equation}
where $\Sigma_d$ is the surface density as a function of radius. Finally, a relation between $\langle v_\phi^2 \rangle$ and the other second moments is needed to fully solve the second order system. Jeans modelling for discs injects the intuition behind a theoretical development known as the \textit{epicycle approximation}.

The main idea is that stars in discs are on roughly circular and planar orbits, so a reasonably accurate description of their orbits can be obtained by taking the equations of motion in a truncated series expansion about a circular orbit at the guiding radius, $R_g$. This is better understood in terms of the effective potential in the rotating reference frame in which a star rotating at $R_g$ is stationary. For a star on the $x$-axis, the $\hat{x}$ direction represents the radial direction, the $\hat{y}$ direction is the same as $\hat{\phi}$ in the original frame, and $\hat{z}$ is unchanged. The equations of motion are \citep{BT},
\begin{eqnarray}
\frac{\text{d}^2x}{\text{d} t^2} &=& \deriv{\Phi_{eff}}{R}\\
\frac{\text{d}^2y}{\text{d} t^2} &=& 0\\
\frac{\text{d}^2z}{\text{d} t^2} &=& \deriv{\Phi_{eff}}{z}
\end{eqnarray}
where $\Phi_{eff} = \Phi(R,z) + L_z^2/2 R^2$ is the effective potential in the accelerating (rotating) reference frame, and $L_z$ is the $z$-component of the angular momentum. These equations can be written in a first order expansion as,
\begin{eqnarray}
\frac{\text{d}^2x}{\text{d} t^2} &=& \kappa^2 x\\
\frac{\text{d}^2y}{\text{d} t^2} &=& 0\\
\frac{\text{d}^2z}{\text{d} t^2} &=& \nu^2 z
\end{eqnarray}
where,
\begin{equation}
\kappa^2(R_g) = \left(R \deriv{\Omega^2}{R} + 4 \Omega^2\right)_{R=R_g,z=0},
\end{equation}
with the circular frequency $\Omega$,
\begin{equation}
\Omega^2(R_g) = \frac{1}{R_g} \left(\deriv{\Phi}{R}\right)_{R=R_g,z=0} = \frac{L_z^2}{R_g^4},
\end{equation}
and,
\begin{equation}
\nu^2(R_g) = \left(\deriv{\Phi}{z}\right)_{R=R_g,z=0}.
\end{equation}
That is, a star's orbit is described by three frequencies: its rotational frequency $\Omega$, a radial oscillation frequency, $\kappa$, and a vertical oscillation frequency, $\nu$. This view is accurate so long as the vertical motion does not deviate far from the disc midplane. The actual values of these constants in Galactic astronomy are related to a problem known as the \textit{Oort Problem}, and we believe that for the Sun \citep{BT},
\begin{eqnarray}
\frac{\kappa}{\Omega} &\approx& 1.35 \\
\frac{\nu}{\kappa} &\approx& 2
\end{eqnarray} 
The fact that the orbit is described by three distinct frequencies that are unequal for the Sun means that the Sun's orbit does not close. If the Milky Way potential never changed, the path of the Sun would fill an annulus centered around $R_g$ \citep{BT}. That is to say, the orbit looks like it forms a rosette pattern.

@@

Although Jeans modelling is used to set up initial conditions for galaxies because it does not require a DF to be known ahead of time, the best initial conditions are generated from DFs \footnote{That is not to say that no one uses Jeans modelling to set up ICs. Approaches like the one presented in \citet{hernquist_1993} were enormously successful and still used today.}. In general, the equilibrium DF is a function of six variables. However, there are some quantities, called \textit{integrals of motion}, which do not change along a star's orbit in a potential. For an axisymmetric potential, these are the total energy, $\mathcal{E}$, $L_z$ and a third quantity, $I_3$.  
For spherical systems, up to five integrals of motion are admitted by the system, but the simplest \textit{ergodic} models simply have $f_{h}(\mathcal{E})$ for the DF.

If we assume the epicycle approximation, then $I_3 \approx E_z = \Phi(R,0) - \Phi(R,z) + \frac{1}{2} v^2$, where $v$ is the velocity of an orbit \citep{kd95galactics}. The vertical energy, $E_z$, is conserved so long as the disc is razor thin. Jeans showed that a DF that is a function of the integrals of motion describes an equilibrium distribution. More specifically, quoting from \citet{BT},
\begin{theorem}[The Jeans Theorem]
Any steady-state solution of the collisionless Boltzmann equation depends on the phase-space coordinates only through integrals of motion in a given potential, and any function of the integrals yields a steady-state solution of the collisionless Boltzmann equation.
\end{theorem}
Proof of this theorem can be obtained in \citet{jeans_1915} and in Chapter~4.2 of \citet{BT}, although it essentially amounts to observing that the time derivative of a function of time-independent quantities is zero by the chain rule. Now that we know that any DF of the integrals of motion gives a steady state solution, and that using the epicyclic approximation gives us three known integrals of motion, we can ask ourselves what kinds of DFs describe the NFW haloes and exponential disks that are easily obtained from Jeans modelling.
